---
title: "gapminder analysis"
author: "Ayub Hareed"
date: "18/01/2022"
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
    highlight: tango

---


```{r setup, include=FALSE}
# load libraries
library(knitr)
library(reactable)
library(tidyverse)
library(RColorBrewer)
library(plotly)

```

## Read the gapminder csv data:

```{r read file}

my_data <- read.csv('../gapminder_clean.csv') 
# change to a tibble
my_data <- as.tibble(my_data)
reactable(my_data)
?reactable
```

## CO2 emissions in 1962

one of the data points (Kuwait) skews the data and can be taken as an outlier

```{r CO2 emissions}
my_data_1962 <- filter(my_data,my_data$Year == 1962)
  
ggplot(data = my_data_1962,aes(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.)) +
geom_point() +
labs(title = 'CO2 emissions vs GDP')


```

Kuwait datasets

```{r pressure}
my_data_1962 %>% 
  filter(my_data_1962$gdpPercap > 50000)

```


## scatter plot without Kuwait data


```{r }
my_data_1962 %>% 
  filter(my_data_1962$gdpPercap < 50000) %>%
  ggplot(aes(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.)) +
  geom_point() +
  labs(title = 'CO2 emissions vs GDP')

```


## Calculate Pearson's R value 

. The **r** value is: 0.9260817
. The **p** value is: less than 2.2e-16

```{r correlation test, echo=FALSE}

x <- as.matrix(my_data_1962[,'gdpPercap'])
y <- as.matrix(my_data_1962[,'CO2.emissions..metric.tons.per.capita.'])
dim(x)
dim(y)
cor.test(x = x,y = y)
```

## Person correletion without Kuwait data:
The correlation coefficient is lower when the Kuwait data 

. The **r** value is: 0.8063295
. The **p** value is: less than 2.2e-16


```{r , echo=FALSE}
my_data_1962_without_kuwait <- filter(my_data_1962,gdpPercap < 50000)
dim(my_data_1962)
dim(my_data_1962_without_kuwait)

x_without_kuwait <- as.matrix(my_data_1962_without_kuwait[,'gdpPercap'])
y_without_kuwait <- as.matrix(my_data_1962_without_kuwait[,'CO2.emissions..metric.tons.per.capita.'])
dim(x_without_kuwait)
dim(y_without_kuwait)
cor.test(x = x_without_kuwait,y = y_without_kuwait)
```

## Calculate correlation for each year:
Note: NA values are removed before calculation for mean values

The year 1967 has the strongest correlation with an r value of 0.9387918

```{r }
my_data_grouped <- my_data %>% group_by(Year)
my_data_grouped %>% 
  summarise(
  R_value = cor.test(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.,)$estimate,
  P_value = cor.test(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.,)$p.value
)


```




## Filter dataset to only include 1967 data

```{r }
my_data_1967 <- my_data %>% filter(Year == 1967)
my_data_1967
```




## ploting a plotly scatter plot:

```{r }
g = ggplot(data = my_data_1967, aes(x = gdpPercap, 
                                y = CO2.emissions..metric.tons.per.capita.,
                                name = Country.Name, 
                                )) +
  geom_point(aes(color = continent,
                 size = pop, 
                 alpha = 0.5)) 
ggplotly(g)

```

# Part two

## 1.  continent vs Energy use:
**box plot** that shows the variations between the continents

```{r }
my_data_continent_grouped <-
  my_data %>% group_by(continent) %>% # group by continent
  drop_na() %>%     # remove missing values
  select(continent,
         Energy.use..kg.of.oil.equivalent.per.capita.) # don't include the missing continent
df_continent_grouped = as.data.frame(my_data_continent_grouped)
dim(df_continent_grouped)
# plot box plot
boxplot(Energy.use..kg.of.oil.equivalent.per.capita. ~ continent, data = df_continent_grouped)
```


### Carry out an Anova test to see if  the difference between the continent is significant.

The tests p-Value is less than <2e-16 showing that their is a significant different between the continent

```{r}
model.aov <- aov(Energy.use..kg.of.oil.equivalent.per.capita. ~ continent, data = df_continent_grouped)
summary(model.aov)
model.aov$coefficients
```



## 2. difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)'

Using an Anova test between Asia and Europe

```{r }
my_data_Asia_Europe <- 
  my_data %>% 
  filter(Year >= 1990 & continent == 'Asia' | continent == 'Europe') %>%
  drop_na() %>%
  select(continent,Year,
         Imports.of.goods.and.services....of.GDP.)
boxplot(Imports.of.goods.and.services....of.GDP. ~ continent, data = my_data_Asia_Europe)
```


**Anova test**
```{r anova}
model.aov2 <- aov(Imports.of.goods.and.services....of.GDP. ~ continent, data = my_data_Asia_Europe)
summary(model.aov2)
```
There anova results shows that there is a significant difference between
Europe and Asia with respect to 'Imports of goods and services' in the years 
after 1990

## Life expactancy after 1962:
1. Create a data frame that contains **country name**, **year** and **life expectancy**.
1. Then group them by country 

```{r }
my_data_life_expactancy <- 
  my_data %>%
  select(Country.Name,Year,Life.expectancy.at.birth..total..years.)
# drop missing values
my_data_life_expactancy <- my_data_life_expactancy %>%
  drop_na() %>%
  group_by(Country.Name)


my_data_life_expactancy

```

```{r }
ggplot(data = my_data_life_expactancy,
       aes(x= Year,
           y= Life.expectancy.at.birth..total..years. )) +
  geom_point(aes(color = Country.Name), show.legend = FALSE)
  
```


## Calculate the total precentage change for each country:

To do this first the year-on-year change in the seven year gap is calculated.


```{r }
# calculate the percentage change over the years
df_life_exp <- 
  my_data_life_expactancy %>%
  mutate(pct_change = (Life.expectancy.at.birth..total..years./lag(Life.expectancy.at.birth..total..years.) -1 )*100) %>%
  summarise(
    total_change = sum(pct_change,na.rm = TRUE)
  )
 
df_life_exp %>% arrange(desc(total_change)) %>%
  slice_head(n=5)
```  
## plot the highest five and lowest five countries in total growth 

```{r}
df_life_exp_upper <- 
  df_life_exp %>% arrange(desc(total_change)) %>%
  slice_head(n=5)

df_life_exp_lower <- 
  df_life_exp %>% arrange(desc(total_change)) %>%
  slice_tail(n=5)
df_life_exp_highest_n_lowest <- bind_rows(df_life_exp_upper,df_life_exp_lower)

country_name_vec <- as.vector(df_life_exp_highest_n_lowest$Country.Name)

# length(country_name.vec)
df_life_exp_highest_n_lowest_full <- my_data_life_expactancy[my_data_life_expactancy$Country.Name %in% country_name_vec,]


ggplot(data = df_life_exp_highest_n_lowest_full, aes(x = Year, y = Life.expectancy.at.birth..total..years.)) +
  geom_point(aes(color = Country.Name)) +
  geom_line(aes(color = Country.Name)) +
  ggtitle( 'Life expectancy from 1960')
  
```

According to this data Cambodia have under gone the highest rate of growth in 
life expectancy.

** plot of year on year change of the same plot

```{r }

df_life_exp_year_on_year <- 
  my_data_life_expactancy %>%
  mutate(pct_change = (Life.expectancy.at.birth..total..years./lag(Life.expectancy.at.birth..total..years.) -1 )*100)



df_life_exp_year_on_year[my_data_life_expactancy$Country.Name %in% country_name_vec,] %>%
  ggplot(aes(x= Year, y = pct_change)) +
  geom_point(aes(color = Country.Name)) + 
  geom_line(aes(color = Country.Name))

barplot(total_change ~ Country.Name, data = df_life_exp_highest_n_lowest, col= brewer.pal(10,"Set3"))

```





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
