---
title: "gapminder analysis"
author: "Ayub Hareed"
date: "18/01/2022"
output: html_document
---

```{r setup, include=FALSE}
# load libraries

library(tidyverse)
```

## Read the gapminder csv data:

```{r read file}

my_data <- read.csv('../gapminder_clean.csv') 
# change to a tibble
my_data <- as.tibble(my_data)
head(my_data)
```

## CO2 emissions in 1962

one of the data points (Kuwait) skews the data and can be taken as an outlier

```{r CO2 emissions}
my_data_1962 <- filter(my_data,my_data$Year == 1962)
  
ggplot(data = my_data_1962,aes(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.)) +
geom_point() +
labs(title = 'CO2 emissions vs GDP')


```

Kuwait datasets

```{r pressure}
my_data_1962 %>% 
  filter(my_data_1962$gdpPercap > 50000)

```


## scatter plot without Kuwait data


```{r }
my_data_1962 %>% 
  filter(my_data_1962$gdpPercap < 50000) %>%
  ggplot(aes(x = gdpPercap,y = CO2.emissions..metric.tons.per.capita.)) +
  geom_point() +
  labs(title = 'CO2 emissions vs GDP')

```


## Calculate Pearson's R value 

. The **r** value is: 0.9260817
. The **p** value is: less than 2.2e-16

```{r pressure, echo=FALSE}

x <- as.matrix(my_data_1962[,'gdpPercap'])
y <- as.matrix(my_data_1962[,'CO2.emissions..metric.tons.per.capita.'])
dim(x)
dim(y)
cor.test(x = x,y = y)
```

## Person correletion without Kuwait data:
The correlation coefficient is lower when the Kuwait data 

. The **r** value is: 0.8063295
. The **p** value is: less than 2.2e-16


```{r pressure, echo=FALSE}
my_data_1962_without_kuwait <- filter(my_data_1962,gdpPercap < 50000)
dim(my_data_1962)
dim(my_data_1962_without_kuwait)

x_without_kuwait <- as.matrix(my_data_1962_without_kuwait[,'gdpPercap'])
y_without_kuwait <- as.matrix(my_data_1962_without_kuwait[,'CO2.emissions..metric.tons.per.capita.'])
dim(x_without_kuwait)
dim(y_without_kuwait)
cor.test(x = x_without_kuwait,y = y_without_kuwait)
```

## Calculate correlation for each year:

```{r }
my_data_grouped <- my_data %>% 

```


## Calculate

```{r }


```


## Calculate correlation for each year:

```{r }


```





## 


```{r pressure, echo=FALSE}
?cor.test
?drop_na
%>%
  cor.test()



gdp_V_co2 <- my_data_1962 %>% 
  drop_na(gdpPercap,CO2.emissions..metric.tons.per.capita.) %>%
  select(gdpPercap,CO2.emissions..metric.tons.per.capita.) 
x <- as.matrix(gdp_V_co2[,1])
y <- as.matrix(gdp_V_co2[,2])
dim(x)
dim(y)
cor.test(x = x,y = y)



```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
